### 2023-02-25 appendices/appendix_benchmarking_workshop.tex
Add appendices for benchmarking and hierarchical

\chapter[benchmarking generative latent variable models for speech]{Benchmarking Generative Latent Variable Models for Speech}

@@

\paragraph{LSTM}
The LSTM baseline uses an MLP encoder to embed the waveform subsegment $\xb_{t:t+s-1}$ to a feature vector before feeding it to the LSTM cell. The encoder is similar to the parameterization of $\phi_\text{vrnn}^\text{enc}$ for the VRNN described above. The LSTM cell produces the hidden state $\db_t$ from $\xb_{t:t+s-1}$ and passes it to a decoder. Like the encoder, the decoder is parameterized like $\phi_\text{vrnn}^\text{dec}$ of the VRNN. It outputs the waveform predictions $\xb_{t+s:t+2s-1}$ from the hidden state $\db_t$. The LSTM model uses a single vanilla unidirectional LSTM cell.

@@

\paragraph{CW-VAE} We implement the CW-VAE based on the original work \cite{saxena_clockwork_2021} but with some modifications also briefly described in section~\cref{sec: cwvae on speech}. We replace the encoder/decoder model architectures of the original work with architectures designed for waveform modeling. Specifically, the encoder and decoder are based on the Conv-TasNet \cite{luo_conv-tasnet_2019} and uses similar residual block structure. However, contrary to the Conv-TasNet, we require downsampling factors larger than two. In order to achieve this we use strides of two in the separable convolution of each block. With e.g. six blocks we hence get an overall stride of $2^6=64$. We can then add additional blocks with unit stride.
We also need to modify the residual connections that skip strided convolutions. Specifically, we replace the residual with a single convolution with stride equal to the stride used in the separable convolution. This convolution uses no nonlinearity and hence simply learns a local linear downsampling.

@@

\paragraph{ASR model}
The ASR model used for the phoneme recognition experiments is a three-layered bidirectional LSTM. We apply temporal dropout between the LSTM layers and also after the final layer. Temporal dropout works similar to regular dropout but samples the entries of the hidden state to mask only once and apply it to all timesteps, i.e. masking $\hb_{t}$ at vector index $i$ for all $t$ (and $i$). 
We mask by zeroing vector elements. We never mask the first timestep. 
We apply temporal dropout with masking probability of 0.3 for the 3.7h subset, 0.35 for the 1h subset and 0.4 for the 10m subset. 
The only difference in model architecture between the evaluation of different representations is the first affine transformation; from the dimensionality of the representation to the hidden state size of the LSTM. This gives rise to a very small difference in model capacity and parameter count which we find is negligible. We set the hidden unit size to 256.

@@

\section{Converting the likelihood to units of bits per frame}\label{app: likelihood in bits per frame}
Here we briefly describe how to compute a likelihood in units of bits per frame (bpf). In the main text, we use $\log$ to mean $\log_e$, but here we will be explicit. In general, conversion from nats to bits (i.e., from $\log_e$ to $\log_2$) is achieved by $\log_2(x) = \log_e(x)  / \log_2(e) $. Remember that $\log_2 p(\xb_{1:T})$ generally factorizes as $\sum_t \log_2 p(x_t|\cdot)$. In sequence modeling, it is important to remember that each example $\xb^{i}$ must be weighted differently according the sequence length of that specific example. This is in contrast to computing bits per dimension in the image domain where images in a dataset are usually of the same dimensions. Thus, we compute the log-likelihood in bits per frame over the entire dataset as
\begin{equation}
    \mathcal{L}(\xb^{i}) = \frac{1}{\sum_i T_i} \sum_i \sum_t \log_2 p(\xb_t^{i}) \enspace ,
\end{equation}
where $i$ denotes the example index, $T_i$ is the length of example $\xb^i$ in waveform frames and $t$ is the time index. If a single timestep $\xb_t^i$ represents multiple waveform frames stacked with some stack size $s$, it is important to note that the sum over $t$ only has $T_i/s$ elements. 
For the LVMs, the term $\log_2 p(\xb_t^{i})$ is lower bounded by the ELBO in \eqref{eq: elbo}.

@@

@@

We constrain the variance of the Gaussians used with our models to be at least $\sigma^2_\text{min} = 0.01^2$ in order to avoid the variance going to zero, the likelihood going to infinity and optimization becoming unstable. The Gaussian standard deviation is clamped at minimum $0.001$ by \cite{aksan_stcn_2019}.

@@

We specifically hypothesize that models that are autoregressive in the observed variable (VRNN, SRNN, Stochastic WaveNet, STCN) are well-equipped to utilize local smoothness to put very high density on the correct next value and that this in turn leads to a high degree of exploitation of the unboundedness of the likelihood. Not being autoregressive in the observed variable, the CW-VAE cannot exploit this local smoothness in the same way. Instead, the reconstruction is conditioned on a stochastic latent variable, $p(\xb_t|\zb^1_t)$, which introduces uncertainty and likely larger reconstruction variances.

\begin{table}[t!]
    \centering
    \begin{tabular}{ll|lr}
        \toprule
        $s$    & \bfseries Model           & \bfseries Configuration           & \bfseries $\mathcal{L}$ [bpf] \\
        \midrule
        1         & Uniform             & Uninformed            & 16.00 \\
        1         & DMoL                & Optimal               & 10.70 \\   % $mean[0.00], scale=[0.005]$
        -         & FLAC                & Linear PCM            & 8.582 \\
        \midrule
        $1$       & Wavenet             & $D_C=96$              & \textbf{7.246} \\
        $1$       & LSTM                & $D_d=256, L=1$        & 7.295 \\
        $1$       & VRNN                & $D_z=256$             & $\leq$7.316 \\
        $1$       & SRNN                & $D_z=256$             & $\leq$7.501 \\
        $1$       & STCN                & $D_z=256,L=5$         & $\leq$9.970 \\
        \midrule
        $64$      & WaveNet             & $D_c=96$              & 8.402 \\
        $64$      & LSTM                & $D_d=256, L=1$        & 8.357 \\
        $64$      & VRNN                & $D_z=256$             & $\leq$8.103 \\
        $64$      & SRNN                & $D_z=256$             & $\leq$8.036 \\
        $64$      & CW-VAE              & $D_z=96, L=1$         & $\leq$7.989 \\
        % $64$      & CW-VAE              & $D_z=96, L=2$         & $\leq$XX \\
        % $64$      & CW-VAE              & $D_z=96, L=3$         & $\leq$XX \\
        $64$      & STCN                & $D_z=256,L=5$  & $\leq$\textbf{7.768} \\
        \midrule
        $256$      & WaveNet             & $D_c=96$              & 9.018 \\
        $256$      & LSTM                & $D_d=256, L=1$        & 8.959 \\  % overfits slightly
        $256$      & VRNN                & $D_z=256$             & $\leq$8.739 \\
        $256$      & SRNN                & $D_z=256$             & $\leq$8.674 \\
        $256$      & CW-VAE              & $D_z=96, L=1$         & $\leq$8.406 \\
        % $256$      & CW-VAE              & $D_z=96, L=2$         & $\leq$XX \\
        % $256$      & CW-VAE              & $D_z=96, L=3$         & $\leq$XX \\
        $256$      & STCN                & $D_z=256,L=5$  & $\leq$\textbf{8.196} \\
        \bottomrule
    \end{tabular}
    \vspace{2mm}
    \caption{
    Model likelihoods on \textbf{TIMIT} represented as a \textbf{16 bit linear PCM}. The STCN converges to a poor local minimum and sometimes diverges when modeling linear PCM with $s=1$.
    }
    \label{tab: timit likelihoods dmol linear appendix}
\end{table}

@@

    \begin{tabular}{ll|lr}

@@

        $s$    & \bfseries Model           & \bfseries Configuration           & \bfseries $\mathcal{L}$ [bpf] \\

@@

    \vspace{2mm}

@@

    Model likelihoods on \textbf{TIMIT} represented as a \textbf{16 bit $\boldsymbol{\mu}$-law encoded PCM}.

@@

\begin{table}[t!]
    \centering
    \begin{tabular}{ll|lr}
        \toprule
        $s$ & \bfseries Model           & \bfseries Configuration           & \bfseries $\mathcal{L}$ [nats] \\
        \midrule
        1 & WaveNet                                                               & Normal & 119656 \\
        1 & WaveNet                                                               & GMM-2  & 120699 \\
        1 & WaveNet                                                               & GMM-20 & 121681 \\
        \midrule
        % 200 & WaveNet                                                              & GMM    & XX.XX
        200 & WaveNet {\scriptsize \cite{aksan_stcn_2019}}                        & GMM-20    & 30188 \\
        200 & WaveNet {\scriptsize \cite{aksan_stcn_2019}}                        & Normal & -7443 \\
        200 & Stochastic WaveNet$^*$ {\scriptsize \cite{lai_stochastic_2018}}     & Normal & $\geq$72463\\
        % 200 & VRNN {\scriptsize \cite{chung_recurrent_2015}}                          & & \\
        200 & VRNN {\scriptsize \cite{chung_recurrent_2015}}                      & Normal & $\approx$28982\\
        % 200 & SRNN {\scriptsize \cite{fraccaro_sequential_2016}}                      & & \\
        200 & SRNN {\scriptsize \cite{fraccaro_sequential_2016}}                  & Normal & $\geq$60550 \\
        200 & STCN {\scriptsize \cite{aksan_stcn_2019}}                           & GMM-20 & $\geq$69195\\
        200 & STCN {\scriptsize \cite{aksan_stcn_2019}}                           & Normal & $\geq$64913\\
        200 & STCN-dense {\scriptsize \cite{aksan_stcn_2019}}                     & GMM-20 & $\geq$71386\\
        200 & STCN-dense {\scriptsize \cite{aksan_stcn_2019}}                     & Normal & $\geq$70294\\
        200 & STCN-dense-large {\scriptsize \cite{aksan_stcn_2019}}               & GMM-20 & $\geq$77438\\
        200 & CW-VAE$^*$                                         & $L=1, D_z=96$, Normal & $\geq$41629 \\  % https://wandb.ai/vseq/cwvae/runs/n8nu778t?workspace=user-jaksen
        % 200 & CW-VAE$^*$                                         & $L=2, D_z=96$, Normal & $\geq$2090 \\  % https://wandb.ai/vseq/cwvae/runs/1so36fii?workspace=user-jaksen
        \bottomrule
    \end{tabular}
    \vspace{2mm}
    \caption{
    Model likelihoods on \textbf{TIMIT} represented as \textbf{globally normalized 16 bit linear PCM}. Contrary to the other likelihoods reported in this paper, here they are given in units of nats and obtained by summing the likelihood over time and over all examples in the dataset and dividing by the total number of examples. In the table, Normal refers to using a Gaussian likelihood and GMM refers to using a  Gaussian Mixture Model likelihood with 20 components. Models with asterisks $^*$ are our implementations while remaining results are as reported in the referenced work.
    }
    \label{tab: timit likelihoods gaussian appendix}
\end{table}

@@

\begin{table}[t!]
    \centering
    \begin{tabular}{rrr}
    \toprule
    $\sigma_\text{min}$ & $\sigma_\text{min}^2$ & $\max\mathcal{L}$ \\
    \midrule
    $1$      & $1$           &    $-45367$ \\
    $0.5$    & $0.25$        &    $-11146$ \\
    $0.1$    & $0.01$        &     $68307$ \\
    $0.05$   & $0.0025$      &    $102525$ \\
    $0.01$   & $0.0001$      &    $181979$ \\
    $0.005$  & $0.000025$    &    $216198$ \\
    $0.001$  & $0.000001$    &    $295651$ \\
    \bottomrule
    \end{tabular}
    \vspace{2mm}
    \caption{The highest possible Gaussian log-likelihoods (max $\mathcal{L}$) attainable on the TIMIT test set as computed by \eqref{eq: gaussian ll with minimum variance} with different values of the minimum variance $\sigma^2_\text{min}$.}
    \label{tab:timit best gaussian ll}
\end{table}

\section{Additional discussion on the choice of output distribution}\label{app: additional discussion on output distribution}
The DMoL uses a discretization of the continuous logistic distribution to define a mixture model over a discrete random variable. This allows it to parameterize multimodal distributions which can express ambiguity about the value of $\xb_t$.
The model can learn to maximize likelihood by assigning a bit of probability mass to multiple potential values of $\xb_t$.

While this is well-suited for autoregressive modeling, for which the distribution was developed, the potential multimodality poses a challenge for non-autoregressive latent variable models which independently sample multiple neighboring observations at the output.
In fact, if multiple neighboring outputs defined by the subsequence $\xb_{t_1:t_2}$ have multimodal $p(\xb_t|\cdot)$, we risk sampling a subsequence where each neighboring value expresses different potential realities, independently.

@@

\section{Additional graphical models}\label{app: additional graphical models}
In \cref{fig: cwvae cell state graph} we show the graphical model of the recurrent cell of the CW-VAE for a single time step. As noted in \cite{saxena_clockwork_2021}, this cell is very similar to the one of the Recurrent State Space Model (RSSM) \cite{hafner_learning_2019}.
In \cref{fig: cwvae three-layered graphical models (inference and generative)} we show the unrolled graphical models of a three-layered CW-VAE with $k_1=1$ and $c=2$ yielding $k_2=2$ and $k_3=4$. 
We show both the generative and inference models and highlight in blue the parameter sharing between the two models due to top-down inference.
In \cref{fig: stcn graphical models (inference and generative)} we show the graphical models of the STCN \cite{aksan_stcn_2019} at a single timestep. The model has three layers and shares the parameters of the WaveNet encoder between the inference and generative models.
In \cref{fig: vrnn graphical models (inference and generative)} we illustrate the unrolled graphical models of the inference and generative models of the VRNN \cite{chung_recurrent_2015}. We include the deterministic variable $\db_t$ in order to illustrate the difference to other latent variable models.
Likewise, in \cref{fig: srnn graphical models (inference and generative)} we illustrate the unrolled graphical models the SRNN \cite{fraccaro_sequential_2016}.

\begin{figure}[t!]
    \centering
    \def\col{blue}
    \begin{tikzpicture}
        % LATENT STATE SEGMENTATION
        % internal latent state
        \node[det] (h_t_l) {$\db_t^l$};
        \node[latent, below right=1cm of h_t_l] (z_t_l) {$\zb_t^l$};
        \node[latent, right=1.5cm of h_t_l] (s_t_l) {$\vs_t^l$};

@@

        % inputs
        \node[latent,left=0.75cm of h_t_l] (s_tm1_l) {$\vs_{t-1}^l$};
        \node[latent,above=0.75cm of h_t_l] (s_t_lp1) {$\vs_t^{l+1}$};%
        \node[latent,below=1cm of z_t_l] (e_t_l) {$\eb_t^l$};%
        % \node[left=0.5cm of e_t_l] (e_t_l_caption) {($\eb_t^l$ conditioned only in inference)};%

@@

        \edge[blue] {s_t_l} {s_t_l_copy1}
        \edge[blue] {s_t_l} {s_t_l_copy2}
    \end{tikzpicture}
\caption{CW-VAE cell state $\vs_t^l$ update. The cell state is given as $\vs_t^l=(\zb_t^l, \db_t^l)$ where $\db_t^l$ is the deterministic hidden state of a Gated Recurrent Unit \cite{cho_properties_2014}. The vector $\eb_t^l$ is computed from $\xb_t$ by the encoder network which outputs $L$ encodings, one for each latent variable, similar to that of a Ladder VAE \cite{sonderby_ladder_2016}. All blue arrows are shared between generation and inference. The dashed arrow is used only during inference. The solid arrow has unique transformations during inference and generation.}
\label{fig: cwvae cell state graph}
\end{figure}

>>>>

\chapter[appendix for benchmarking generative latent variable models for speech]{Benchmarking Generative Latent Variable Models for Speech}

@@

\paragraph{LSTM}
The LSTM baseline uses an MLP encoder to embed the waveform subsegment $\vx_{t:t+s-1}$ to a feature vector before feeding it to the LSTM cell. The encoder is similar to the parameterization of $\phi_\text{vrnn}^\text{enc}$ for the VRNN described above. The LSTM cell produces the hidden state $\vd_t$ from $\vx_{t:t+s-1}$ and passes it to a decoder. Like the encoder, the decoder is parameterized like $\phi_\text{vrnn}^\text{dec}$ of the VRNN. It outputs the waveform predictions $\vx_{t+s:t+2s-1}$ from the hidden state $\vd_t$. The LSTM model uses a single vanilla unidirectional LSTM cell.

@@

\paragraph{CW-VAE} We implement the CW-VAE based on the original work \cite{saxena_clockwork_2021} but with some modifications also briefly described in the paper. We replace the encoder/decoder model architectures of the original work with architectures designed for waveform modeling. Specifically, the encoder and decoder are based on the Conv-TasNet \cite{luo_conv-tasnet_2019} and uses similar residual block structure. However, contrary to the Conv-TasNet, we require downsampling factors larger than two. In order to achieve this we use strides of two in the separable convolution of each block. With e.g. six blocks we hence get an overall stride of $2^6=64$. We can then add additional blocks with unit stride.
We also need to modify the residual connections that skip strided convolutions. Specifically, we replace the residual with a single convolution with stride equal to the stride used in the separable convolution. This convolution uses no nonlinearity and hence simply learns a local linear downsampling.

@@

\paragraph{ASR model}
The ASR model used for the phoneme recognition experiments is a three-layered bidirectional LSTM. We apply temporal dropout between the LSTM layers and also after the final layer. Temporal dropout works similar to regular dropout but samples the entries of the hidden state to mask only once and apply it to all timesteps, i.e. masking $\vh_{t}$ at vector index $i$ for all $t$ (and $i$). 
We mask by zeroing vector elements. We never mask the first timestep. 
We apply temporal dropout with masking probability of 0.3 for the 3.7h subset, 0.35 for the 1h subset and 0.4 for the 10m subset. 
The only difference in model architecture between the evaluation of different representations is the first affine transformation; from the dimensionality of the representation to the hidden state size of the LSTM. This gives rise to a very small difference in model capacity and parameter count which we find is negligible. We set the hidden unit size to 256.

@@

\section{Converting the likelihood to units of bits per frame}\label{app: likelihood in bits per frame}
Here we briefly describe how to compute a likelihood in units of bits per frame (bpf). In the main text, we use $\log$ to mean $\log_e$, but here we will be explicit. In general, conversion from nats to bits (i.e., from $\log_e$ to $\log_2$) is achieved by $\log_2(x) = \log_e(x)  / \log_2(e) $. Remember that $\log_2 p(\vx_{1:T})$ generally factorizes as $\sum_t \log_2 p(x_t|\cdot)$. In sequence modeling, it is important to remember that each example $\vx^{i}$ must be weighted differently according the sequence length of that specific example. This is in contrast to computing bits per dimension in the image domain where images in a dataset are usually of the same dimensions. Thus, we compute the log-likelihood in bits per frame over the entire dataset as
\begin{equation}
    \mathcal{L}(\vx^{i}) = \frac{1}{\sum_i T_i} \sum_i \sum_t \log_2 p(\vx_t^{i}) \enspace ,
\end{equation}
where $i$ denotes the example index, $T_i$ is the length of example $\vx^i$ in waveform frames and $t$ is the time index. If a single timestep $\vx_t^i$ represents multiple waveform frames stacked with some stack size $s$, it is important to note that the sum over $t$ only has $T_i/s$ elements. 
For the LVMs, the term $\log_2 p(\vx_t^{i})$ is lower bounded by the ELBO in \eqref{eq: elbo}.

@@

\paragraph{LibriSpeech, $\mu$-law, DMoL} We provide additional results on LibriSpeech with audio represented as $\mu$-law encoded PCM in \cref{tab: likelihoods librispeech}. See appendix~\cref{app: dataset details}, \cref{app: model architectures} and \cref{app: training details} for additional details.

@@

We constrain the standard deviation of the Gaussians used with our models to be at least $\sigma_\text{min} = 0.01$ in order to avoid it going to zero, the likelihood going to infinity and optimization becoming unstable. The minimum Gaussian standard deviation of \citet{aksan_stcn_2019} is $\sigma_\text{min} = 0.001$.

@@

We specifically hypothesize that models that are autoregressive in the observed variable (VRNN, SRNN, Stochastic WaveNet, STCN) are well-equipped to utilize local smoothness to put very high density on the correct next value and that this in turn leads to a high degree of exploitation of the unboundedness of the likelihood. Not being autoregressive in the observed variable, the CW-VAE cannot exploit this local smoothness in the same way. Instead, the reconstruction is conditioned on a stochastic latent variable, $p(\vx_t|\vz^1_t)$, which introduces uncertainty and likely larger reconstruction variances.

\begin{table*}[t]
    \centering
    \resizebox{0.98\textwidth}{!}{
    \begin{tabular}{lll|rrrr}  % https://wandb.ai/vseq/icml_asr
        \toprule
        $s$ & \bf Model                   & \bf Configuration       & \multicolumn{4}{c}{\bf Likelihood $\mathcal{L}$ [bpf]} \\
         & & & dev-clean & dev-other & test-clean & test-other\\
         & & & 10h/100h & 10h/100h & 10h/100h & 10h/100h\\
        \midrule
        1 & Uniform             & Uninformed     & \multicolumn{1}{c}{16.00} & \multicolumn{1}{c}{16.00} & \multicolumn{1}{c}{16.00} & \multicolumn{1}{c}{16.00} \\
        1 & DMoL                & Optimal        & \multicolumn{1}{c}{15.66} & \multicolumn{1}{c}{15.70} & \multicolumn{1}{c}{15.62} & \multicolumn{1}{c}{15.71}\\   % $\vmu=[-0.551, 0.551], \vz=[0.11, 0.11]$
        - & FLAC                & Linear PCM     & \multicolumn{1}{c}{\bf 9.390} & \multicolumn{1}{c}{\bf 9.292} & \multicolumn{1}{c}{\bf 9.700} & \multicolumn{1}{c}{\bf 9.272} \\
        \midrule
        1 & Wavenet             & $D_c=96$       & \bf 10.96/10.89 & \bf 10.85/10.76 & \bf 11.12/11.01 & \bf 11.05/10.85 \\
        1 & LSTM                & $D_d=256$      & 11.21/11.17 & 11.10/11.06 & 11.35/11.29 & 11.28/11.23 \\  % 15ua3hau/(lmssrk5k,248n2ylu)
        \midrule
        64 & Wavenet            & $D_c=96$       & 13.61/13.24 & 13.58/13.21 & 13.61/13.22 & 13.60/13.21 \\  % 2ghkybtd/3iieasdt
        64 & LSTM               & $D_d=256$      & 13.56/13.25 & 13.52/13.24 & 13.55/13.23 & 13.56/13.25 \\  % 10fpovu3/2bu1izin
        64 & CW-VAE             & $D_z=96,L=1$   & $\leq$12.32/12.24 & 12.32/12.23 & 12.43/12.33 & 12.43/12.33 \\
        64 & CW-VAE             & $D_z=96,L=2$   & $\leq$12.30/12.22 & 12.30/12.21 & 12.40/12.31 & 12.39/12.32 \\
        64 & STCN               & $D_z=256,L=5$  & $\leq$\bf 11.83/11.47 & \bf 11.82/11.46 & \bf 11.94/11.58 & \bf 11.94/11.60 \\  % kutxm2hx/3t0xg7pb
        \bottomrule
    \end{tabular}
    }
    \caption{
    Model likelihoods $\mathcal{L}$ on \textbf{LibriSpeech} test sets represented as \textbf{16 bit $\boldsymbol{\mu}$-law encoded PCM}.
    For the CW-VAE, $s$ refers to $s_1$ and the two-layered models have $s_2=8s_1$.
    The models are trained on either the $\SI{10}{h}$ LibriLight subset or the $\SI{100}{h}$ LibriSpeech \textrm{train-clean-100} subset as indicated.
    Likelihoods are given in units of bits per frame (bpf).
    }
    \label{tab: likelihoods librispeech}
\end{table*}

\begin{table}[t!]
    \centering
    \begin{tabular}{lll|r}
        \toprule
        $s$    & \bf Model              & \bf Configuration     & \bf $\mathcal{L}$ [bpf] \\
        \midrule
        1         & Uniform             & Uninformed            & 16.00 \\
        1         & DMoL                & Optimal               & 10.70 \\   % $mean[0.00], scale=[0.005]$
        %-         & FLAC                & Linear PCM            & 8.582 \\
        \midrule
        $1$       & Wavenet             & $D_C=96$              & \textbf{7.246} \\
        $1$       & LSTM                & $D_d=256, L=1$        & 7.295 \\
        $1$       & VRNN                & $D_z=256$             & $\leq$7.316 \\
        $1$       & SRNN                & $D_z=256$             & $\leq$7.501 \\
        $1$       & STCN                & $D_z=256,L=5$         & $\leq$9.970 \\
        \midrule
        $64$      & WaveNet             & $D_c=96$              & 8.402 \\
        $64$      & LSTM                & $D_d=256, L=1$        & 8.357 \\
        $64$      & VRNN                & $D_z=256$             & $\leq$8.103 \\
        $64$      & SRNN                & $D_z=256$             & $\leq$8.036 \\
        $64$      & CW-VAE              & $D_z=96, L=1$         & $\leq$7.989 \\
        % $64$      & CW-VAE              & $D_z=96, L=2$         & $\leq$XX \\
        % $64$      & CW-VAE              & $D_z=96, L=3$         & $\leq$XX \\
        $64$      & STCN                & $D_z=256,L=5$  & $\leq$\textbf{7.768} \\
        \midrule
        $256$      & WaveNet             & $D_c=96$              & 9.018 \\
        $256$      & LSTM                & $D_d=256, L=1$        & 8.959 \\  % overfits slightly
        $256$      & VRNN                & $D_z=256$             & $\leq$8.739 \\
        $256$      & SRNN                & $D_z=256$             & $\leq$8.674 \\
        $256$      & CW-VAE              & $D_z=96, L=1$         & $\leq$8.406 \\
        % $256$      & CW-VAE              & $D_z=96, L=2$         & $\leq$XX \\
        % $256$      & CW-VAE              & $D_z=96, L=3$         & $\leq$XX \\
        $256$      & STCN                & $D_z=256,L=5$         & $\leq$\textbf{8.196} \\
        \bottomrule
    \end{tabular}
    \caption{
    Model likelihoods on \textbf{TIMIT} represented as a \textbf{16 bit linear PCM}, obtained by different latent variable models and compared to autoregressive baselines all using a discretized mixture of logistics with 10 components as output distribution. Likelihoods are given in units of bits per frame (bpf) and obtained by normalizing the total likelihood of each sequence with the individual sequence length and then averaging over the dataset. The STCN converges to a poor local minimum and sometimes diverges when modeling linear PCM with $s=1$.
    }
    \label{tab: timit likelihoods dmol linear appendix}
\end{table}

@@

    \begin{tabular}{lll|r}

@@

        $s$    & \bf Model           & \bf Configuration           & \bf $\mathcal{L}$ [bpf] \\

@@

@@

    Model likelihoods on \textbf{TIMIT} represented as a \textbf{16 bit $\boldsymbol{\mu}$-law encoded PCM}, obtained by different latent variable models and compared to autoregressive baselines all using a discretized mixture of logistics with 10 components as output distribution. Likelihoods are given in units of bits per frame (bpf) and obtained by normalizing the total likelihood of each sequence with the individual sequence length and then averaging over the dataset.

@@

\begin{table}[t!]
    \centering
    \begin{tabular}{lll|r}
        \toprule
        $s$ & \bf Model           & \bf Configuration           & \bf $\mathcal{L}$ [nats] \\
        \midrule
        1 & WaveNet                                                               & Normal & 119656 \\
        1 & WaveNet                                                               & GMM-2  & 120699 \\
        1 & WaveNet                                                               & GMM-20 & 121681 \\
        \midrule
        % 200 & WaveNet                                                              & GMM    & XX.XX
        200 & WaveNet {\scriptsize \cite{aksan_stcn_2019}}                        & GMM-20    & 30188 \\
        200 & WaveNet {\scriptsize \cite{aksan_stcn_2019}}                        & Normal & -7443 \\
        200 & Stochastic WaveNet$^*$ {\scriptsize \cite{lai_stochastic_2018}}     & Normal & $\geq$72463\\
        % 200 & VRNN {\scriptsize \cite{chung_recurrent_2015}}                          & & \\
        200 & VRNN {\scriptsize \cite{chung_recurrent_2015}}                      & Normal & $\approx$28982\\
        % 200 & SRNN {\scriptsize \cite{fraccaro_sequential_2016}}                      & & \\
        200 & SRNN {\scriptsize \cite{fraccaro_sequential_2016}}                  & Normal & $\geq$60550 \\
        200 & STCN {\scriptsize \cite{aksan_stcn_2019}}                           & GMM-20 & $\geq$69195\\
        200 & STCN {\scriptsize \cite{aksan_stcn_2019}}                           & Normal & $\geq$64913\\
        200 & STCN-dense {\scriptsize \cite{aksan_stcn_2019}}                     & GMM-20 & $\geq$71386\\
        200 & STCN-dense {\scriptsize \cite{aksan_stcn_2019}}                     & Normal & $\geq$70294\\
        200 & STCN-dense-large {\scriptsize \cite{aksan_stcn_2019}}               & GMM-20 & $\geq$77438\\
        200 & CW-VAE$^*$                                         & $L=1, D_z=96$, Normal & $\geq$41629 \\  % https://wandb.ai/vseq/cwvae/runs/n8nu778t?workspace=user-jaksen
        % 200 & CW-VAE$^*$                                         & $L=2, D_z=96$, Normal & $\geq$2090 \\  % https://wandb.ai/vseq/cwvae/runs/1so36fii?workspace=user-jaksen
        \bottomrule
    \end{tabular}
    \caption{
    Model likelihoods on \textbf{TIMIT} represented as \textbf{globally normalized 16 bit linear PCM}. Likelihoods are given in units of nats and obtained by summing the likelihood over time and over all examples in the dataset and dividing by the total number of examples. In the table, Normal refers to using a Gaussian likelihood and GMM refers to using a  Gaussian Mixture Model likelihood with 20 components. Models with asterisks $^*$ are our implementations while remaining results are as reported in the referenced work.
    }
    \label{tab: timit likelihoods gaussian appendix}
\end{table}

@@

\begin{table}[t!]
    \centering
    \begin{tabular}{rrr}
    \toprule
    $\sigma_\text{min}$ & $\sigma_\text{min}^2$ & $\max\mathcal{L}$ \\
    \midrule
    $1$      & $1$           &    $-45367$ \\
    $0.5$    & $0.25$        &    $-11146$ \\
    $0.1$    & $0.01$        &     $68307$ \\
    $0.05$   & $0.0025$      &    $102525$ \\
    $0.01$   & $0.0001$      &    $181979$ \\
    $0.005$  & $0.000025$    &    $216198$ \\
    $0.001$  & $0.000001$    &    $295651$ \\
    \bottomrule
    \end{tabular}
    \caption{The highest possible Gaussian log-likelihoods (max $\mathcal{L}$) attainable on the TIMIT test set as computed by \eqref{eq: gaussian ll with minimum variance} with different values of the minimum variance $\sigma^2_\text{min}$.}
    \label{tab:timit best gaussian ll}
\end{table}

\section{Additional discussion on the choice of output distribution}\label{app: additional discussion on output distribution}
The DMoL uses a discretization of the continuous logistic distribution to define a mixture model over a discrete random variable. This allows it to parameterize multimodal distributions which can express ambiguity about the value of $\vx_t$.
The model can learn to maximize likelihood by assigning a bit of probability mass to multiple potential values of $\vx_t$.

While this is well-suited for autoregressive modeling, for which the distribution was developed, the potential multimodality poses a challenge for non-autoregressive latent variable models which independently sample multiple neighboring observations at the output.
In fact, if multiple neighboring outputs defined by the subsequence $\vx_{t_1:t_2}$ have multimodal $p(\vx_t|\cdot)$, we risk sampling a subsequence where each neighboring value expresses different potential realities, independently.

@@

\section{Additional graphical models}\label{app: additional graphical models}
In \cref{fig: cwvae cell state graph} we show the graphical model of the recurrent cell of the CW-VAE for a single time step. As noted in \cite{saxena_clockwork_2021}, this cell is very similar to the one of the Recurrent State Space Model (RSSM) \cite{hafner_learning_2019}.

In \cref{fig: cwvae three-layered graphical models (inference and generative)} we show the unrolled graphical models of a three-layered CW-VAE with $k_1=1$ and $c=2$ yielding $k_2=2$ and $k_3=4$. 
We show both the generative and inference models and highlight in blue the parameter sharing between the two models due to top-down inference.

In \cref{fig: stcn graphical models (inference and generative)} we show the graphical models of the STCN \cite{aksan_stcn_2019} at a single timestep. The model has three layers and shares the parameters of the WaveNet encoder between the inference and generative models.

In \cref{fig: vrnn graphical models (inference and generative)} we illustrate the unrolled graphical models of the inference and generative models of the VRNN \cite{chung_recurrent_2015}. We include the deterministic variable $\vd_t$ in order to illustrate the difference to other latent variable models.

Likewise, in \cref{fig: srnn graphical models (inference and generative)} we illustrate the unrolled graphical models the SRNN \cite{fraccaro_sequential_2016}.

\begin{figure}[t!]
    \centering
    \def\col{blue}
    \begin{tikzpicture}
        % LATENT STATE SEGMENTATION
        % internal latent state
        \node[det] (h_t_l) {$\vd_t^l$};
        \node[latent, below right=1cm of h_t_l] (z_t_l) {$\vz_t^l$};
        \node[latent, right=1.5cm of h_t_l] (s_t_l) {$\vs_t^l$};

@@

        % inputs
        \node[latent,left=0.75cm of h_t_l] (s_tm1_l) {$\vs_{t-1}^l$};
        \node[latent,above=0.75cm of h_t_l] (s_t_lp1) {$\vs_t^{l+1}$};%
        \node[latent,below=1cm of z_t_l] (e_t_l) {$\ve_t^l$};%
        % \node[left=0.5cm of e_t_l] (e_t_l_caption) {($\ve_t^l$ conditioned only in inference)};%

@@

        \edge[blue] {s_t_l} {s_t_l_copy1}
        \edge[blue] {s_t_l} {s_t_l_copy2}
    \end{tikzpicture}
\caption{CW-VAE cell state $\vs_t^l$ update. The cell state is given as $\vs_t^l=(\vz_t^l, \vd_t^l)$ where $\vd_t^l$ is the deterministic hidden state of a Gated Recurrent Unit \cite{cho_properties_2014}. The vector $\ve_t^l$ is computed from $\vx_t$ by the encoder network which outputs $L$ encodings, one for each latent variable, similar to that of a Ladder VAE \cite{sonderby_ladder_2016}. All blue arrows are shared between generation and inference. The dashed arrow is used only during inference. The solid arrow has unique transformations during inference and generation.}
\label{fig: cwvae cell state graph}
\end{figure}

\begin{figure}[t!]
    \begin{subfigure}[b]{\textwidth}
    \flushright
    \def\col{blue}
    \resizebox{!}{5.25cm}{
    \begin{tikzpicture}
        % GENERATIVE
        % nodes
        \node[obs] (x_1) {$\vx_t$};%
        \node[obs,right=1.00cm of x_1] (x_2) {$\vx_{t+1}$};%
        \node[obs,right=1.00cm of x_2] (x_3) {$\vx_{t+2}$};%
        \node[obs,right=1.00cm of x_3] (x_4) {$\vx_{t+3}$};%
        \node[obs,right=1.00cm of x_4] (x_5) {$\vx_{t+4}$};%
        \node[obs,right=1.00cm of x_5] (x_6) {$\vx_{t+5}$};%
        \node[obs,right=1.00cm of x_6] (x_7) {$\vx_{t+6}$};%
        \node[obs,right=1.00cm of x_7] (x_8) {$\vx_{t+7}$};%
